---
title: "Analysis of cytochrome b mutational spectra in fishes"
author: "Alina G. Mikhailova"
date: "`r Sys.Date()`"
output:
    workflowr::wflow_html:
    toc: true
knitr:
  opts_chunk:
    echo: false
---

```{r setup, echo=FALSE, include=FALSE}
docname <- "I_CYTB_polymorphic_mutspec_analysis"
now <- Sys.time()
# Time chunks during knitting
knitr::knit_hooks$set(timeit = function(before) {
  if (before) {
    print(paste("Start:", Sys.time()))
    now <<- Sys.time()
  } else {
    print(paste("Stop:", Sys.time()))
    print(Sys.time() - now)
  }
})

knitr::opts_chunk$set(
  autodep        = TRUE,
  cache          = FALSE,
  cache.path     = paste0("cache/", docname, "/"),
  cache.comments = FALSE,
  cache.lazy     = FALSE,
  dev            = c("png", "pdf"),
  echo           = TRUE,
  error          = FALSE,
  fig.align      = "center",
  fig.width      = 14,
  fig.height     = 12,
  message        = FALSE,
  warning        = FALSE,
  timeit         = TRUE
)
# Load packages
suppressPackageStartupMessages({
  library(here)
  library(knitr)
  library(tidyverse)
  library(magrittr)
  library(caper)
  library(geiger)
  library(ggExtra)
  library(ggpubr)
  library(broom)
  library(phytools)
  library(ggstatsplot)
  library(rfishbase)
})

# Set paths
src_dir <- here("code")
data_dir <- here("data")
output_dir <- here("output")
plots_dir <- here(output_dir, "figures")
tables_dir <- here(output_dir, "tables")

# parallelisation
n_cores <- 4

# set seed
reseed <- 13
set.seed(seed = reseed)
```

# Reading mutation spectra database (L-strand) and ecological databases

```{r cytb_spectra, message=FALSE}
mutSpec12 <- read_csv(here(data_dir, "MutSpecVertebrates12fish.csv"))
mutSpec12Cb <- dplyr::filter(mutSpec12, Gene == "Cytb")
cat("N species =", length(unique(mutSpec12Cb$Species)))
kable(head(mutSpec12Cb))

MutSpecForAnalysisCb <- mutSpec12Cb %>% dplyr::select(Class, Species, Mut, MutSpec)
MutSpecForAnalysisCb$Mut <- gsub(">", "_", MutSpecForAnalysisCb$Mut)
MutSpecForAnalysisCb <- MutSpecForAnalysisCb %>% pivot_wider(names_from = Mut, values_from = MutSpec, , values_fn = median) 
colnames(MutSpecForAnalysisCb) <- c("Class",     # Only CytB
                                  "Species", 
                                  "T_G.H", 
                                  "T_C.H", 
                                  "T_A.H", 
                                  "G_T.H", 
                                  "G_C.H", 
                                  "G_A.H", 
                                  "C_T.H", 
                                  "C_G.H", 
                                  "C_A.H", 
                                  "A_T.H", 
                                  "A_G.H", 
                                  "A_C.H")
```

```{r both_genes_spectra}
MutSpecForAnalysisBoth <- mutSpec12 %>% dplyr::select(Class, Species, Mut, MutSpec)
MutSpecForAnalysisBoth$Mut <- gsub(">", "_", MutSpecForAnalysisBoth$Mut)
MutSpecForAnalysisBoth <- MutSpecForAnalysisBoth %>% pivot_wider(names_from = Mut, values_from = MutSpec, , values_fn = median) 
cat("N species =", length(unique(mutSpec12$Species)))
colnames(MutSpecForAnalysisBoth) <- c("Class",      # Both genes
                                  "Species", 
                                  "T_G.H", 
                                  "T_C.H", 
                                  "T_A.H", 
                                  "G_T.H", 
                                  "G_C.H", 
                                  "G_A.H", 
                                  "C_T.H", 
                                  "C_G.H", 
                                  "C_A.H", 
                                  "A_T.H", 
                                  "A_G.H", 
                                  "A_C.H")
```


```{r fish_ecology}
fishEcology <- read.csv(here(data_dir, "Full_fish_ecology_table_2025.csv"))
kable(head(fishEcology))
cor.test(fishEcology$tm, fishEcology$Temp_avg, method = "spearman")
```

# Calculating of average MutSpec for Fisches
## Cytb

```{r average_mutspec_plot_cytb, message=FALSE}
AverageMutSpecCb <- mutSpec12Cb %>% group_by(Mut) %>% summarise(meanmutspec = mean(MutSpec, na.rm = TRUE))
AverageMutSpecCb$Mut.H <- c("T_G.H", 
                          "T_C.H", 
                          "T_A.H", 
                          "G_T.H", 
                          "G_C.H", 
                          "G_A.H", 
                          "C_T.H", 
                          "C_G.H", 
                          "C_A.H", 
                          "A_T.H", 
                          "A_G.H", 
                          "A_C.H")

f1 = ggplot(AverageMutSpecCb,
  aes(
    x = AverageMutSpecCb$Mut.H,
    y = AverageMutSpecCb$meanmutspec,
    fill = AverageMutSpecCb$Mut.H
    )
  ) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c(
      "#bdbdbd",
      "#00a426",
      "#bdbdbd",
      "#bdbdbd",
      "#bdbdbd",
      "red",
      "#900000",
      "#bdbdbd",
      "#bdbdbd",
      "#bdbdbd",
      "darkgreen",
      "#bdbdbd"
    )
  ) +
  labs(title = "Average mitochondrial mutational spectrum in Fishes (cytB, N=840)", x = "Substitution types", y = "Normalised Frequencies") + theme(
    legend.position = "none",
    panel.background = element_blank(),
    axis.line = element_line(color = "black")
  )

ggsave(here(plots_dir,"Average_mutspec_in_Fishes_CytB.pdf"),
  plot = f1,
  width = 6,
  height = 6
)
f1
```

## Both genes

```{r average_mutspec_plot_both, message=FALSE}
AverageMutSpecBoth <- mutSpec12 %>% group_by(Mut) %>% summarise(meanmutspec = mean(MutSpec, na.rm = TRUE))
AverageMutSpecBoth$Mut.H <- c("T_G.H", 
                          "T_C.H", 
                          "T_A.H", 
                          "G_T.H", 
                          "G_C.H", 
                          "G_A.H", 
                          "C_T.H", 
                          "C_G.H", 
                          "C_A.H", 
                          "A_T.H", 
                          "A_G.H", 
                          "A_C.H")

f1 = ggplot(AverageMutSpecBoth,
  aes(
    x = AverageMutSpecBoth$Mut.H,
    y = AverageMutSpecBoth$meanmutspec,
    fill = AverageMutSpecBoth$Mut.H
    )
  ) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c(
      "#bdbdbd",
      "#00a426",
      "#bdbdbd",
      "#bdbdbd",
      "#bdbdbd",
      "#ff0000",
      "#900000",
      "#bdbdbd",
      "#bdbdbd",
      "#bdbdbd",
      "#005213",
      "#bdbdbd"
    )
  ) +
  labs(title = "Average mitochondrial mutational spectrum in Fishes (cytB+COX1, N=1035)", x = "Substitution types", y = "Normalised Frequencies") + theme(
    legend.position = "none",
    # Remove legend
    panel.background = element_blank(),
    axis.line = element_line(color = "black")# Remove background
  )

f1
```

# Comparison of average MutSpec between cold and warm fisches
```{r average_mutspec_cytb_groups}
merging <- list(MutSpecForAnalysisCb, fishEcology)
MutAndEcoCb <- merging %>% purrr::reduce(left_join, by= "Species")

#write.table(MutAndEcoCb, file = here(tables_dir, "Supplementary_Material_1_Mitochondrial_mutational_spectra_cytb_and_ecology_Fishes.txt"), row.names = FALSE)
#write_csv(MutAndEcoCb, file = here(tables_dir, "Supplementary_Material_1_Mitochondrial_mutational_spectra_cytb_and_ecology_Fishes.csv"))

temperaturestat <- summary(MutAndEcoCb$Temp_avg)

WarmFishesCb <- filter(MutAndEcoCb, Temp_avg > as.numeric(temperaturestat["3rd Qu."]))
WarmFishesCb <- WarmFishesCb %>% dplyr::select(Species, T_G.H, 
                                        T_C.H, 
                                        T_A.H, 
                                        G_T.H, 
                                        G_C.H, 
                                        G_A.H, 
                                        C_T.H, 
                                        C_G.H, 
                                        C_A.H, 
                                        A_T.H, 
                                        A_G.H, 
                                        A_C.H)
WarmFishesCb <- pivot_longer(WarmFishesCb, cols = c("T_G.H", 
                                  "T_C.H", 
                                  "T_A.H", 
                                  "G_T.H", 
                                  "G_C.H", 
                                  "G_A.H", 
                                  "C_T.H", 
                                  "C_G.H", 
                                  "C_A.H", 
                                  "A_T.H", 
                                  "A_G.H", 
                                  "A_C.H"), names_to = "Mut", values_to = "MutSpec" )
NwarmCb <- length(unique(WarmFishesCb$Species))

ColdFishesCb <- filter(MutAndEcoCb, Temp_avg < as.numeric(temperaturestat["1st Qu."]))
ColdFishesCb <- ColdFishesCb %>% dplyr::select(Species, T_G.H, 
                                        T_C.H, 
                                        T_A.H, 
                                        G_T.H, 
                                        G_C.H, 
                                        G_A.H, 
                                        C_T.H, 
                                        C_G.H, 
                                        C_A.H, 
                                        A_T.H, 
                                        A_G.H, 
                                        A_C.H)
ColdFishesCb <- pivot_longer(ColdFishesCb, cols = c("T_G.H", 
                                  "T_C.H", 
                                  "T_A.H", 
                                  "G_T.H", 
                                  "G_C.H", 
                                  "G_A.H", 
                                  "C_T.H", 
                                  "C_G.H", 
                                  "C_A.H", 
                                  "A_T.H", 
                                  "A_G.H", 
                                  "A_C.H"), names_to = "Mut", values_to = "MutSpec" )
NcoldCb <- length(unique(ColdFishesCb$Species))

f1a <- ggplot(
  WarmFishesCb,
  aes(
    x = Mut,
    y = MutSpec
  )
) +
  geom_jitter(aes(colour = Mut), width = 0.1) +
  scale_color_manual(
    values = c(
      "#bdbdbd",
      "#00a426",
      "#bdbdbd",
      "#bdbdbd",
      "#bdbdbd",
      "red",
      "#900000",
      "#bdbdbd",
      "#bdbdbd",
      "#bdbdbd",
      "darkgreen",
      "#bdbdbd"
    )
  ) +
  labs(
    title = "Warm Fishes, >23C",
    x = "Substitution types",
    y = "Normalised Frequencies"
  ) +
  theme(
    legend.position = "none",
    panel.background = element_blank(),
    axis.line = element_line(color = "black")
  ) +
  geom_text(
    x = 2.5,
    y = 0.63,
    label = paste("N warm =", NwarmCb),
    size = 7
  ) +
  stat_summary(
    fun = "median",
    geom = "line",        
    aes(group = 2),      
    color = "black",     
    size = 0.5             
  )


f1b <- ggplot(
 ColdFishesCb,
  aes(
    x = Mut,
    y = MutSpec
  )
) +
  geom_jitter(aes(colour = Mut), width = 0.1) +
  scale_color_manual(
    values = c(
      "#bdbdbd",
      "#00a426",
      "#bdbdbd",
      "#bdbdbd",
      "#bdbdbd",
      "red",
      "#900000",
      "#bdbdbd",
      "#bdbdbd",
      "#bdbdbd",
      "darkgreen",
      "#bdbdbd"
    )
  ) +
  labs(
    title = "Cold Fishes, <13C",
    x = "Substitution types",
    y = "Normalised Frequencies"
  ) +
  theme(
    legend.position = "none",
    panel.background = element_blank(),
    axis.line = element_line(color = "black")
  ) +
  geom_text(
    x = 2.5,
    y = 0.56,
    label = paste("N cold =", NcoldCb),
    size = 7
  ) +
  stat_summary(
    fun = "median",
    geom = "line",        
    aes(group = 2),      
    color = "black",     
    size = 0.5             
  )

f1 <- ggarrange(f1b, f1a, ncol = 2)

ggsave(here(plots_dir,"Average mitochondrial mutational spectrum in Fishes divided by groups (cytB).pdf"),
  plot = f1,
  width = 12,
  height = 6
)
f1
```

# Studies of A_G spectra with temperature and other parameters in fishes

## A_G freqs dependence on temperature

```{r statistics_with_temp_cytb, message = FALSE, warning = FALSE} 
correlation_results <- MutAndEcoCb |>
  dplyr::select(
    "A_G.H",
    "A_T.H",
    "A_C.H",
    "T_A.H",
    "T_G.H",
    "T_C.H",
    "C_A.H",
    "C_T.H",
    "C_G.H",
    "G_T.H",
    "G_A.H",
    "G_C.H"
  ) |>
  map_dfr( ~ tidy(cor.test(.x, MutAndEcoCb$Temp_avg), method = 'spearman'), .id = "variable") |>
  arrange(desc(abs(estimate)))

kable(correlation_results)
#write.table(kable(correlation_results), file = here(tables_dir, "Table1.txt"), row.names = FALSE, col.names = FALSE)

SubsetClass<- unique(MutAndEcoCb$Class)
spN <- nrow(MutAndEcoCb[!is.na(MutAndEcoCb$Temp_avg),])
cat("N species =", spN, "Taxa =", SubsetClass)

f1 <- ggplot(MutAndEcoCb, aes(x = Temp_avg, y = A_G.H)) +
  geom_point(color = "#73514f") +                  # Scatter plot
  geom_smooth(
    method = "lm",
    color = "black",
    # Regression line + CI
    fill = "lightgray",
    se = TRUE
  ) +     # Set se=TRUE for confidence interval
  stat_cor(
    method = "spearman",
    # Spearman correlation stats
    aes(label = paste(..r.label.., ..p.label.., "N", spN, sep = "~"))
  ) +  # Include n
  labs(title = "Positive assosiation of A>G.H with temperature", x = "Median annual water temperature, ?C", y = "AH>GH") +
  theme(
    legend.position = "none",
    panel.background = element_blank(),
    axis.line = element_line(color = "black")
  )
 
ggsave(here(plots_dir,"Assosiation_of_A_G_with_temperature.pdf"),
  plot = f1,
  width = 6,
  height = 6
) 
f1
```


## A_G freqs dependence on climate zones


```{r comparison_climate_A_G}
ClimateZonetest <- na.omit(data.frame(subset(MutAndEcoCb, select = c(Species ,A_G.H, Climate_zone))))

ClimateZonetest$ClimateGroup <- ifelse(ClimateZonetest$Climate_zone %in% c("tropical", "subtropical"), "Warm", "Cold")
ClimateZonetest$Climate_zone <- factor(ClimateZonetest$Climate_zone, levels = c("tropical", "subtropical", "temperate", "boreal", "polar"), ordered = TRUE)
AGmedian <- median(ClimateZonetest$A_G.H)
ClimateZonetest$A_G.Freq <- "low_A_G"
ClimateZonetest[ClimateZonetest$A_G.H >= AGmedian,]$A_G.Freq <- "high_A_G"
ClimateZonetest$A_G.Freq <- as.factor(ClimateZonetest$A_G.Freq)
ClimateZonePie <- ClimateZonetest[,2-3]
ClimPl <- c("#F2AF4AFF", "#EB7F54FF", "#C36377FF", "#61599DFF", "#1D457FFF")
names(ClimPl) <- c("tropical", "subtropical", "temperate", "boreal", "polar") 
s1 <- ggpiestats(data = ClimateZonePie, x = Climate_zone, y = A_G.Freq) +
  scale_fill_manual(values = ClimPl)
s1

tidy(wilcox.test(ClimateZonetest[ClimateZonetest$ClimateGroup == "Warm",]$A_G.H, ClimateZonetest[ClimateZonetest$ClimateGroup == "Cold",]$A_G.H))

```

## A\>G freqs dependence on time of maturation

```{r maturation_cytb, message = FALSE, warning = FALSE}
correlation_results1 <- MutAndEcoCb |>
  dplyr::select(
    "A_G.H",
    "A_T.H",
    "A_C.H",
    "T_A.H",
    "T_G.H",
    "T_C.H",
    "C_A.H",
    "C_T.H",
    "C_G.H",
    "G_T.H",
    "G_A.H",
    "G_C.H"
  ) |>
  map_dfr( ~ tidy(cor.test(.x, MutAndEcoCb$tm), method = 'spearman'), .id = "variable") |>
  arrange(desc(abs(estimate)))
kable(correlation_results1)
```

## PGLS

```{r}
create_enhanced_summary_table <- function(model_summary, pgls_model) {
  # Extract coefficients
  coef_table <- model_summary$coefficients
  
  # Create enhanced dataframe
  enhanced_df <- data.frame(
    Parameter = rownames(coef_table),
    Estimate = round(coef_table[, "Estimate"], 6),
    Std_Error = round(coef_table[, "Std. Error"], 6),
    t_value = round(coef_table[, "t value"], 3),
    p_value = format.pval(coef_table[, "Pr(>|t|)"], digits = 3),
    Significance = case_when(
      coef_table[, "Pr(>|t|)"] < 0.001 ~ "***",
      coef_table[, "Pr(>|t|)"] < 0.01 ~ "**",
      coef_table[, "Pr(>|t|)"] < 0.05 ~ "*",
      coef_table[, "Pr(>|t|)"] < 0.1 ~ ".",
      TRUE ~ ""
    ),
    CI_Lower = round(coef_table[, "Estimate"] - 1.96 * coef_table[, "Std. Error"], 6),
    CI_Upper = round(coef_table[, "Estimate"] + 1.96 * coef_table[, "Std. Error"], 6)
  )
  
  # Model fit statistics
  model_stats <- data.frame(
    Statistic = c(
      "R-squared",
      "Adjusted R-squared",
      "F-statistic",
      "Lambda",
      "AIC",
      "BIC"
    ),
    Value = c(
      round(model_summary$r.squared, 4),
      round(model_summary$adj.r.squared, 4),
      round(model_summary$fstatistic[1], 3),
      round(pgls_model$param["lambda"], 6),
      round(AIC(pgls_model), 2),
      round(BIC(pgls_model), 2)
    )
  )
  
  return(list(coefficients = enhanced_df, model_stats = model_stats))
}
```

```{r inertia_analysis_temperature, message=FALSE, warning=FALSE}
tree <- read.tree(file = here(data_dir, "chordates_species.nwk"))
tree <- ape::drop.tip(tree, "Tenebrosternarchus_preto")
MutAndEcoPGLS <- MutAndEcoCb %>% dplyr::select(Species, A_G.H, C_T.H, G_A.H, T_C.H, Temp_avg)

row.names(MutAndEcoPGLS) = MutAndEcoPGLS$Species
tree_pruned = treedata(tree, MutAndEcoPGLS, sort=T, warnings=T)$phy
data<-as.data.frame(treedata(tree_pruned, MutAndEcoPGLS, sort=T, warnings=T)$data)
data$Species = as.character(data$Species)
data$A_G.H = as.numeric(data$A_G.H)
data$C_T.H = as.numeric(data$C_T.H)
data$G_A.H = as.numeric(data$G_A.H)
data$T_C.H = as.numeric(data$T_C.H)
data$Temp_avg = as.numeric(data$Temp_avg)
data$Tv <- 1-(data$A_G.H + data$G_A.H + data$T_C.H + data$C_T.H)
data_comp <- comparative.data(tree_pruned, data, Species, vcv=TRUE)

mod <- pgls(A_G.H ~ Temp_avg, data_comp, lambda="ML")
summ <- summary( pgls(A_G.H ~ Temp_avg, data_comp, lambda="ML"))


pgls_results <- create_enhanced_summary_table(summ, mod)

kable(pgls_results$coefficients)
kable(pgls_results$model)

#write.table(kable(pgls_results$coefficients), file = here(tables_dir, "Table2a.txt"), row.names = FALSE, col.names = FALSE)
#write.table(kable(pgls_results$model), file = here(tables_dir, "Table2b.txt"), row.names = FALSE, col.names = FALSE)
```

```{r inertia_analysis_climate, message=FALSE, warning=FALSE}
tree <- read.tree(file = here(data_dir, "chordates_species.nwk"))
tree <- ape::drop.tip(tree, "Tenebrosternarchus_preto")

row.names(ClimateZonetest) = ClimateZonetest$Species
tree_pruned = treedata(tree, ClimateZonetest, sort=T, warnings=T)$phy
data<-as.data.frame(treedata(tree_pruned, ClimateZonetest, sort=T, warnings=T)$data)
data$Species = as.character(data$Species)
data$A_G.H = as.numeric(data$A_G.H)
data$ClimateGroup = as.factor(data$ClimateGroup)
data_comp <- comparative.data(tree_pruned, data, Species, vcv=TRUE)



mod <- pgls(A_G.H ~ ClimateGroup, data_comp, lambda="ML")
summ <- summary(pgls(A_G.H ~ ClimateGroup, data_comp, lambda="ML"))


pgls_results <- create_enhanced_summary_table(summ, mod)

kable(pgls_results$coefficients)
kable(pgls_results$model)

```

```{r lambda, eval=FALSE}
la.A_G <- phylosig(tree_pruned, data$A_G.H, method = "lambda", test = TRUE) #Phylogenetic signal lambda : 0.30935 , P-value (based on LR test) : 3.47115e-05
la.C_T <- phylosig(tree_pruned, data$C_T.H, method = "lambda", test = TRUE) #Phylogenetic signal lambda : 0.216484, P-value (based on LR test) : 0.000556152 
la.G_A <- phylosig(tree_pruned, data$G_A.H, method = "lambda", test = TRUE) #Phylogenetic signal lambda : 0.522227, P-value (based on LR test) : 5.35018e-05 
la.T_C <- phylosig(tree_pruned, data$T_C.H, method = "lambda", test = TRUE) #Phylogenetic signal lambda : 0.853065 
la.Tv <- phylosig(tree_pruned, data$Tv, method = "lambda", test = TRUE)     #Phylogenetic signal lambda : 0.138137 
                                                                            #####################################

signal_strength <- data.frame(
  Level = c("None", "Weak", "Moderate", "Strong"),
  Range = c("0-0.15", "0.15-0.35", "0.35-0.75", "0.75-1.0")
)
signal_strength$Level <- factor(
  signal_strength$Level,
  levels = c("None", "Weak", "Moderate", "Strong"),
  ordered = TRUE
)

f1a <- ggplot(signal_strength, aes(x=2, y=Level, fill=Level)) +
  geom_col(alpha=0.8, width=1) +
  geom_point(aes(x=1, y=2), size=15, color="#e74c3c") + 
  annotate("text", x=1, y=2, label=paste("λ =", la.A_G$lambda), size=5, fontface="bold") +
  scale_fill_manual(values=c("#ecf0f1", "#bdc3c7", "#3498db", "#2c3e50")) +
  labs(title="Phylogenetic Signal Strength of A_G.H", ) +
  theme_void() +
  theme(legend.position = "none")

f1b <- ggplot(signal_strength, aes(x=2, y=Level, fill=Level)) +
  geom_col(alpha=0.8, width=1) +
  geom_point(aes(x=1, y=2), size=15, color="#e74c3c") + 
  annotate("text", x=1, y=2, label=paste("λ =", la.C_T$lambda), size=5, fontface="bold") +
  scale_fill_manual(values=c("#ecf0f1", "#bdc3c7", "#3498db", "#2c3e50")) +
  labs(title="Phylogenetic Signal Strength of C_T.H") +
  theme_void()

f1c <- ggplot(signal_strength, aes(x=2, y=Level, fill=Level)) +
  geom_col(alpha=0.8, width=1) +
  geom_point(aes(x=1, y=3), size=15, color="#e74c3c") + 
  annotate("text", x=1, y=3, label=paste("λ =", la.G_A$lambda), size=5, fontface="bold") +
  scale_fill_manual(values=c("#ecf0f1", "#bdc3c7", "#3498db", "#2c3e50")) +
  labs(title="Phylogenetic Signal Strength of G_A.H") +
  theme_void() +
  theme(legend.position = "none")

f1d <- ggplot(signal_strength, aes(x=2, y=Level, fill=Level)) +
  geom_col(alpha=0.8, width=1) +
  geom_point(aes(x=1, y=4), size=15, color="#e74c3c") + 
  annotate("text", x=1, y=4, label=paste("λ =", la.T_C$lambda), size=5, fontface="bold") +
  scale_fill_manual(values=c("#ecf0f1", "#bdc3c7", "#3498db", "#2c3e50")) +
  labs(title="Phylogenetic Signal Strength of T_C.H") +
  theme_void() +
  theme(legend.position = "none")

fig <- (f1a | f1b) / (f1c | f1d) 

MutAndEcoPGLS <- MutAndEcoPGLS[!is.na(MutAndEcoPGLS$Temp_avg),]
row.names(MutAndEcoPGLS) = MutAndEcoPGLS$Species
tree_pruned = treedata(tree, MutAndEcoPGLS, sort=T, warnings=T)$phy
data<-as.data.frame(treedata(tree_pruned, MutAndEcoPGLS, sort=T, warnings=T)$data)
data$Species = as.character(data$Species)
data$A_G.H = as.numeric(data$A_G.H)
data$C_T.H = as.numeric(data$C_T.H)
data$G_A.H = as.numeric(data$G_A.H)
data$T_C.H = as.numeric(data$T_C.H)
data$Temp_avg = as.numeric(data$Temp_avg)
data$Tv <- 1-(data$A_G.H + data$G_A.H + data$T_C.H + data$C_T.H)

la.temp <- phylosig(tree_pruned, data$Temp_avg, method = "lambda", test = TRUE) ########Phylogenetic signal lambda : 0.921383, P-value (based on LR test) : 8.81583e-12 


f1e <- ggplot(signal_strength, aes(x=2, y=Level, fill=Level)) +
  geom_col(alpha=0.8, width=1) +
  geom_point(aes(x=1, y=4), size=15, color="#e74c3c") + # Position at "Moderate"
  annotate("text", x=1, y=4, label=paste("λ =", la.temp$lambda), size=7, fontface="bold") +
  scale_fill_manual(values=c("#ecf0f1", "#bdc3c7", "#3498db", "#2c3e50")) +
  labs(title="Phylogenetic Signal Strength of temperature") +
  theme_void() +
  theme(legend.position = "none")

f1f <- ggplot(signal_strength, aes(x=2, y=Level, fill=Level)) +
  geom_col(alpha=0.8, width=1) +
  geom_point(aes(x=1, y=1), size=20, color="#e74c3c") + # Position at "Moderate"
  annotate("text", x=1, y=1, label=paste("λ =", la.Tv$lambda), size=7, fontface="bold") +
  scale_fill_manual(values=c("#ecf0f1", "#bdc3c7", "#3498db", "#2c3e50")) +
  labs(title="Phylogenetic Signal Strength of transversions") +
  theme_void() 

figf <- (f1e | f1f)

ggsave(here(plots_dir,"Lambda_temperature_and_Tv.pdf"),
  plot = figf,
  width = 9,
  height = 9
) 

ggsave(here(plots_dir,"Lambda_transitions.pdf"),
  plot = fig,
  width = 8,
  height = 8
) 

create_lambda_summary_table <- function(la_result, varia) {
  values <- c("variable" = varia,
    `Lambda (λ) estimate` = la_result$lambda,
    `Log-likelihood (λ)` = la_result$logL,
    `Likelihood Ratio (λ=0)` = la_result$logL0,
    `P-value (LR test)` = la_result$P
  )
  df_wide <- as.data.frame(t(values), check.names = FALSE)
  rownames(df_wide) <- NULL
  return(df_wide)
}

AG <- create_lambda_summary_table(la.A_G, "A>G.H")
TC <- create_lambda_summary_table(la.T_C, "T>C.H")
GA <- create_lambda_summary_table(la.G_A, "G>A.H")
CT <- create_lambda_summary_table(la.C_T, "C>T.H") 

lambda_results <- bind_rows(AG, TC, GA, CT, .id = NULL)

#write.table(kable(lambda_results), file = here(tables_dir, "Table3.txt"), row.names = FALSE, col.names = FALSE)

```